Reasoning About a Machine with Local Capabilities

### 2018-01-18 esop-2018/paper.tex
Fixed some of the todos with Lars. Citations and attacker model still on todo.

%% Title information
\title{Reasoning About a Capability Machine with Local Capabilities}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Provably Safe Stack and Return Pointer Management (without OS Support)}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

\begin{abstract}
  Capability machines provide security guarantees at machine level which makes
  them an interesting target for compilation schemes that provably enforce
  properties like control-flow correctness and encapsulation of local state. We
  provide a formalization of a representative capability machine with local
  capabilities and study a novel calling convention for enforcing control-flow
  correctness and encapsulation of local state on it. To prove these properties,
  we provide a logical relation that semantically captures the guarantees
  provided by the hardware (a form of capability safety). The logical relation
  is not specific to our calling convention and can be used to reason about
  arbitrary programs.
\end{abstract}

@@

\textbf{Program start-up} We assume that the language runtime initializes the
memory as follows: a contiguous array of memory is reserved for the stack, for
which we receive a stack pointer in a special register $r_\stk$. We stress that
the stack is not built-in, but merely an abstraction we put on this piece of the
memory. The stack pointer is local and has $\rwlx$ permission. Crucially, the
stack is the only part of memory for which the runtime (including malloc,
loading, linking) will ever provide $\rwlx$ or $\rwl$ capabilities.
Additionally, our examples typically also assume some memory to store
instructions or static data. Another part of memory (called the heap) is
initially governed by malloc and at program start-up, no other code has
capabilities for this memory. Malloc hands out $\rwx$ capabilities for allocated
regions as requested (no $\rwlx$ or $\rwl$ permissions). For simplicity, we
assume that memory allocated through malloc cannot be freed.

@@

\textbf{Reusable macro instructions} We define a number of reusable macros
capturing the calling convention and other conveniences. All macros that use the
stack assume a stack pointer in register $r_\stk$. The macro
\texttt{\footnotesize{fetch $r$ $\var{name}$}} fetches the capability related to
$\var{name}$ from the linking table and stores it in register $r$.  The macros
\texttt{\footnotesize{push $r$}} and \texttt{\footnotesize{pop $r$}} add and
remove elements from the stack. The macro \texttt{\footnotesize{prepstk $r$}} is
used when a callback is invoked by the adversary and prepares the received stack
pointer by checking that it has permission $\rwlx$.  The macro
\texttt{\footnotesize{scall
    $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)}} jumps to the
capability in register $r$ in the manner described above. That is, it pushes
local state (the contents of registers $\overline{r_{\var{priv}}}$) and the
activation record (return code, return pointer, stack pointer) to the stack,
creates an $\entry$ return pointer, restricts the stack pointer, clears the
unused part of the stack, clears the necessary registers and jumps to $r$.  Upon
return, the private state is restored. The macro \texttt{\footnotesize{mclear
    $r$}} clears all the memory the capability in register $r$ has authority
over.  The macro \texttt{\footnotesize{rclear $\var{regSet}$}} clears all the
registers in $\var{regSet}$. The macro \texttt{\footnotesize{reqglob $r$}}
checks whether the word in register $r$ is a global capability. The macro
\texttt{\footnotesize{crtcls $\overline{(x_i,r_i)}$ $r$}} allocates a closure
where $r$ points to the closure's code and a new environment is allocated (using
malloc) where the contents of $\overline{r_i}$ is stored. In the code referred
to by $r$, an implicit fetch happens when an instruction refers to $x_i$.

@@

\section{Examples}
\label{sec:examples}
In this section, we demonstrate how our formalization of capability
safety allows us to prove local-state encapsulation and control-flow
correctness properties for challenging program examples. The security
measures of \sectionname~\ref{sec:stack-and-return-pointer} are deployed to
ensure these properties. Since we are dealing with assembly language,
there are many details to the formal treatment, and therefore we
necessarily omit some details in the lemma statements. The
examples may look deceivingly short, but it is because they use the
previously described macro instructions, which expands to multiple
basic instructions. The examples would be unintelligible without the
macros. The interested reader can find all the technical details in
the technical appendix~\cite{technical_appendix}.

@@

>>>>

%% Title information
%\title{Reasoning About a Capability Machine with Local Capabilities}         %% [Short Title] is optional;
\title{Reasoning About a Machine with Local Capabilities}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Provably Safe Stack and Return Pointer Management (without OS Support)}                     %% \subtitle is optional
\subtitle{Provably Safe Stack and Return Pointer Management}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

\begin{abstract}
  Capability machines provide security guarantees at machine level which makes
  them an interesting target for secure compilation schemes that provably
  enforce properties such as control-flow correctness and encapsulation of local
  state. We provide a formalization of a representative capability machine with
  local capabilities and study a novel calling convention.  We provide a logical
  relation that semantically captures the guarantees provided by the hardware (a
  form of capability safety) and use it to prove control-flow correctness and
  encapsulation of local state.  The logical relation is not specific to our
  calling convention and can be used to reason about arbitrary programs.
\end{abstract}

@@

\textbf{Program start-up} We assume that the language runtime initializes the
memory as follows: a contiguous array of memory is reserved for the stack, for
which we receive a stack pointer in a special register $r_\stk$. We stress that
the stack is not built-in, but merely an abstraction we put on this piece of the
memory. The stack pointer is local and has $\rwlx$ permission. Note that this
means that we will be placing and executing instructions on the stack.
Crucially, the stack is the only part of memory for which the runtime (including
malloc, loading, linking) will ever provide $\rwlx$ or $\rwl$ capabilities.
Additionally, our examples typically also assume some memory to store
instructions or static data. Another part of memory (called the heap) is
initially governed by malloc and at program start-up, no other code has
capabilities for this memory. Malloc hands out $\rwx$ capabilities for allocated
regions as requested (no $\rwlx$ or $\rwl$ permissions). For simplicity, we
assume that memory allocated through malloc cannot be freed.

@@

\textbf{Reusable macro instructions} 
\label{para:macros}
We define a number of reusable macros
capturing the calling convention and other conveniences. All macros that use the
stack assume a stack pointer in register $r_\stk$. The macro
\texttt{\footnotesize{fetch $r$ $\var{name}$}} fetches the capability related to
$\var{name}$ from the linking table and stores it in register $r$.  The macros
\texttt{\footnotesize{push $r$}} and \texttt{\footnotesize{pop $r$}} add and
remove elements from the stack. The macro \texttt{\footnotesize{prepstk $r$}} is
used when a callback is invoked by the adversary and prepares the received stack
pointer by checking that it has permission $\rwlx$.  The macro
\texttt{\footnotesize{scall
    $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)}} jumps to the
capability in register $r$ in the manner described above. That is, it pushes
local state (the contents of registers $\overline{r_{\var{priv}}}$) and the
activation record (return code, return pointer, stack pointer) to the stack,
creates an $\entry$ return pointer, restricts the stack pointer, clears the
unused part of the stack, clears the necessary registers and jumps to $r$.  Upon
return, the private state is restored. The macro \texttt{\footnotesize{mclear
    $r$}} clears all the memory the capability in register $r$ has authority
over.  The macro \texttt{\footnotesize{rclear $\var{regSet}$}} clears all the
registers in $\var{regSet}$. The macro \texttt{\footnotesize{reqglob $r$}}
checks whether the word in register $r$ is a global capability. The macro
\texttt{\footnotesize{crtcls $\overline{(x_i,r_i)}$ $r$}} allocates a closure
where $r$ points to the closure's code and a new environment is allocated (using
malloc) where the contents of $\overline{r_i}$ is stored. In the code referred
to by $r$, an implicit fetch happens when an instruction refers to $x_i$.

@@

\section{Examples}
\label{sec:examples}
In this section, we demonstrate how our formalization of capability safety
allows us to prove local-state encapsulation and control-flow correctness
properties for challenging program examples. The security measures
of \sectionname~\ref{sec:stack-and-return-pointer} are deployed to ensure these
properties. Since we are dealing with assembly language, there are many details
to the formal treatment, and therefore we necessarily omit some details in the
lemma statements. The examples may look deceivingly short, but it is because
they use the macro instructions described in Section \ref{para:macros}.
The examples would be
unintelligible without the macros, as each macro expands to multiple basic instructions. The interested reader can find all the
technical details in the technical appendix~\cite{technical_appendix}.

@@

\section*{Acknowledgements}
This research was supported in part by the ModuRes Sapere Aude Advanced Grant
from The Danish Council for Independent Research for the Natural Sciences (FNU).
